---
title: "Further Exploration and Model"
subtitle: "4/30 Presentation"
format: 
  html:
    page-layout: full
    theme: cosmo
    toc: true
    toc-title: "Contents"
    embed-resources: true
    standalone: true
editor: 
  markdown: 
    wrap: 72
---

# Further Exploration

We noticed from EDA that the `count` variable was very right-skewed, and had a large range including regions with very sparse data. To examine this further, and to potentially determine if there was a range of the `count` data that could be used for prediction, we decided to look at the range from a count of 10 to the mean (51.9) plus one standard deviation (1616.5), which is 1668.4 birds. 
```{r}
#| label: loading in the data
#| echo: true
#| warning: false
#| message: false 
library(tidyverse)

#read in the cleaned data
data <- read_csv("birds_clean.csv")

# Plot the censored data
data_smallcount <- data %>%
  filter(count > 10, count < 1668.4)

ggplot(data_smallcount, aes(x = count)) + 
  geom_histogram(binwidth = 10) +
  labs(
    title = "Bird Count: Censored (10 < count < 1668.4)",
    x = "Bird Count"
  ) +
  theme_minimal()
```

Looking at this much smaller region of the `count` data compared to the whole dataset, this is much more interpretable, but we can still see the issue of a heavy right-skew and regions with sparse data. We then decided to categorize the data into quantiles indicating a scale of number of birds from no birds to very many birds. 

```{r}
#| label: quantiled bird count data
#| echo: true
#| warning: false
#| message: false 

# Maybe use Pandas for this

```

Since our `count` data was not Normal, we decided to do some ensemble methods and more flexible modeling. 