---
title: "Birds"
author: "Sydney Kahn and Jacqueline Adelmann"
format: html
editor: visual
---

# Exploring the data 

We are looking at bird sightings at sea, which comes from TidyTuesday at https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-04-14. Each observation communicates the data that an observer witnessed during a 10-minute period at a port. 

```{r}
#| label: loading in the data
#| echo: true
#| warning: false
#| message: false 

tuesdata <- tidytuesdayR::tt_load('2026-04-14')

beaufort_scale <- tuesdata$beaufort_scale
birds <- tuesdata$birds
sea_states <- tuesdata$sea_states
ships <- tuesdata$ships

```

```{r}
#| label: load packages
#| echo: true 
#| warning: false
#| message: false

library(tidyverse) # for general purposes and ggplot 
library(gtsummary) # for nice tables
```

```{r}
#| label: merging dfs
#| echo: true

# merge ship and sea data 
ships <- ships %>%
  left_join(beaufort_scale, by = "wind_speed_class") %>%
  left_join(sea_states, by = "sea_state_class") %>%
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date))

# merge birds with ship and sea data
df <- birds %>%
  left_join(ships, by = "record_id")
```

```{r}
#| label: managing missing data
#| include: false

# total number of missing values
sum(is.na(df))
# proportion of missing values in entire dataset
mean(is.na(df))

#want to see if certain cols have a lot missing
missing_summary_raw_df <- df %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_raw_df

# 
# common_name (or group species!)
# protocol_type
# duration_minutes
# effort_distance_km
# number_observers
# hour
# season
# location (stat
#           
#clean the data. did it this way because there is a missing data problem both by row and by column, and clearing either individually would leave us with no data 

df_clean <- df %>%
  filter(!is.na(count)) %>% # these are variables we definitely want to use, so filter out observations that are missing these. species common name has the fewest missing obvs of the three different types of species name columns. doing them in separate lines because the or command wasn't working correctly for some reasons
  filter(!is.na(latitude)) %>%
  filter(!is.na(longitude)) %>%
  filter(!is.na(year)) %>%
  filter(!is.na(month)) %>%
  filter(!is.na(species_common_name)) %>% 
  select(where(~ mean(is.na(.)) <= 0.6)) %>% 
  filter(rowMeans(is.na(.)) <= 0.6) %>%
  select(where(~ mean(is.na(.)) <= 0.2)) %>%
  filter(rowMeans(is.na(.)) <= 0.0) 

# see where we're at after removing some things 
missing_summary_cleaned_df <- df_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_cleaned_df

# total number of missing values
sum(is.na(df_clean))
# proportion of missing values in entire dataset
mean(is.na(df_clean))

```

```{r}
#| label: making some new variables
#| echo: false

# species column has a LOT of species that are observed very few times. going to make a new variable that collapses rare ones into one "other" column
df_clean <- df_clean %>%
  group_by(species_common_name) %>%
  mutate(species_count = n()) %>%
  ungroup() %>%
  mutate(species_simplified = case_when(
    species_count/nrow(df_clean) >= 0.01 ~ species_common_name,
    TRUE ~ "other"
  )) %>%
  select(-species_count)


# this still has a lot of unique birds. it may be better to do family level
df_clean <- df_clean %>%
  mutate(species_family = case_when(
    str_detect(species_common_name, "albatross|mollymawk") ~ "Albatross",
    str_detect(species_common_name, "petrel|shearwater|prion|fulmar|diving-petrel") ~ "Petrel/Shearwater/Prion",
    str_detect(species_common_name, "gull|tern|noddy") ~ "Gull/Tern",
    str_detect(species_common_name, "skua|jaeger") ~ "Skua/Jaeger",
    str_detect(species_common_name, "penguin") ~ "Penguin",
    str_detect(species_common_name, "gannet|booby") ~ "Gannet/Booby",
    str_detect(species_common_name, "cormorant|shag") ~ "Cormorant/Shag",
    str_detect(species_common_name, "frigatebird|tropicbird") ~ "Frigatebird/Tropicbird",
    TRUE ~ "Unidentified/Other"
  )) 

count(df_clean, species_family)
```

```{r}
#| label: summary stats table 
#| echo: false

# factoring species so that unidentified comes last 
df_clean <- df_clean %>%
  mutate(species_family = factor(species_family,
                                levels = c(
                                  "Albatross",
                                  "Petrel/Shearwater/Prion",
                                  "Gull/Tern",
                                  "Skua/Jaeger",
                                  "Penguin",
                                  "Gannet/Booby",
                                  "Cormorant/Shag",
                                  "Frigatebird/Tropicbird",
                                  "Unidentified/Other"
                                )))

# print table. selecting our core variables, which are the ones that we are investigating in EDA 
df_clean %>%
  select(
    species_family,
    count,
    season,
    cloud_cover, precipitation,
    feeding, flying_past, following_ship,
    sea_state_description,
    longitude,
    latitude
  ) %>%
  mutate(
    season = str_to_title(season),
    precipitation = str_to_title(precipitation),
    cloud_cover = str_to_title(cloud_cover),
    sea_state_description = str_to_title(sea_state_description)
  ) %>%
  rename(
         "Bird Family" = species_family,
         "Bird count" = count,
         "Number of birds feeding" = feeding,
         "Number of birds flying past" = flying_past,
         "Number of birds following ships" = following_ship,
         "Cloud cover" = cloud_cover,
         "Season" = season,
         "Precipitation" = precipitation,
         "Sea state" = sea_state_description,
         "Longitude" = longitude,
         "Latitude" = latitude
         ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
    as_gt() %>%
  gt::tab_caption("Table 1. Summary of bird observations")
```

Our key outcome variable currently is bird count, which measures how many total birds an observer witnessed during an observation session. We are secondarily interested in looking at counts of birds doing specific activities, such as feeding, flying past, and following ships. Our key explanatory variables are season, longitude, latitude, precipitation, cloud cover, and sea state. We expect that the number of birds will vary dramatically depending on the season and location, but that poor weather will also affect bird count at a more granular level. 


# Data Wrangling and Transformation

We wrangled the data via the base R object `tt_load`, which allows us to directly load TidyTuesday data from Github. We merged the separate data frames by the TidyTuesday-recommend identifier columns recommended to form a data frame with `r dim(df)[1]` observations of `r dim(df)[2]` variables. 

As for data cleaning, we chose to exclude some observations and variables with missing data to ensure that our exploratory assessments were not distorted by data that would need to be removed for modeling purposes. We first removed any observations that were missing our core variables: bird count, species common name, latitude, longitude, year, and month. Next, we performed an iterative process of removing columns and observations with missing data; we removed variables with more than 60% of observations missing, then observations with more than 60% of variables missing, then variables with more than 20% of observations missing, and, finally, observations with any missing data. This iterative process allowed us to retain `r dim(df_clean)[1]` observations and `r dim(df_clean)[2]` variables. 

We created a new variable that categorized birds by family rather than species because most individual species were observed very few times, but overall families were observed frequently.

******Add a codebook – a description of all variables you are using, including ones you are creating for this

# Data Visualization
```{r}
#| label: viz-one
#| fig-cap: This figure shows 
#| echo: true
 
# grab summary so we can plot iqr 
df_summary <- df_clean %>%
  distinct(record_id, .keep_all = TRUE) %>% # so we don't count one session multiple times 
  group_by(season) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    q1 = quantile(count, 0.25, na.rm = TRUE),
    q3 = quantile(count, 0.75, na.rm = TRUE)
  ) 

ggplot(df_summary, aes(x = season, y = mean_count)) + 
  geom_col() +
  geom_errorbar(aes(ymin = q1, ymax = q3), width = 0.2) +
  labs(title = "Mean Number of Birds Observed by Season with IQR",
       x = "Season",
       y = "Mean Number of Birds") +
  theme_minimal()

```

```{r}
#| label: viz-two
#| fig-cap: This figure shows xyz
#| echo: true

# figure
df_geo <- df_clean %>%
  distinct(record_id, .keep_all = TRUE) %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  select(longitude, latitude)

ggplot(df_geo, aes(x = longitude, y = latitude)) +
  geom_point(alpha = 0.4)

```

```{r}
#| label: viz-three
#| fig-cap: This figure showssss
#| echo: true

# figure

```

```{r}
#| label: viz-four
#| fig-cap: This figure shows abc
#| echo: true

# figure

```
