---
title: "Birds"
author: "Sydney Kahn and Jacqueline Adelmann"
format: html
editor: visual
---

# Exploring the data 
```{r}
#| label: loading in the data
#| echo: true
#| warning: false
#| message: false 

tuesdata <- tidytuesdayR::tt_load('2026-04-14')

beaufort_scale <- tuesdata$beaufort_scale
birds <- tuesdata$birds
sea_states <- tuesdata$sea_states
ships <- tuesdata$ships

```

```{r}
#| label: load packages
#| echo: true 
#| warning: false
#| message: false

library(tidyverse) # for general purposes and ggplot 
library(kableExtra) # for nice tables
```

```{r}
#| label: merging dfs
#| echo: true

# merge ship and sea data 
ships <- ships %>%
  left_join(beaufort_scale, by = "wind_speed_class") %>%
  left_join(sea_states, by = "sea_state_class") %>%
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date))

# merge birds with ship and sea data
df <- birds %>%
  left_join(ships, by = "record_id")
```

```{r}
#| label: managing missing data
#| include: false

# total number of missing values
sum(is.na(df))
# proportion of missing values in entire dataset
mean(is.na(df))


#want to see if certain cols have a lot missing
missing_summary_raw_df <- df %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_raw_df

#clean the data. did it this way because there is a missing data problem both by row and by column, and clearing either individually would leave us with no data 
df_clean <- df %>%
  select(-sex) %>% # sex is entirely missing
  select(where(~ mean(is.na(.)) <= 0.2)) %>% 
  filter(rowMeans(is.na(.)) <= 0.2)

# see where we're at after removing some things 
missing_summary_cleaned_df <- df_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_cleaned_df

# total number of missing values
sum(is.na(df_clean))
# proportion of missing values in entire dataset
mean(is.na(df_clean))


# this is pretty good. we could still remove some more cols if we deem it necessary, but this might be good enough
```


```{r}
#| label: summary statistics table
#| echo: true

# i don't like how i did this. need to improve 
sessions <- df_clean %>%
  group_by(record_id) %>%
  summarise(
    total_birds = sum(count, na.rm = TRUE),
    n_species = n_distinct(species_common_name),
    .groups = "drop"
  ) %>%
  left_join(ships, by = "record_id")


# need to add aesthetic items 
kable(sessions)

```

We are looking at bird sightings at sea, which comes from TidyTuesday at https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-04-14.

Answer the following questions:
What is your outcome variable(s)? How well does it measure the outcome you are interested? How does it
relate to your expectations?
What are your key explanatory variables?
In addition, create a table of summary statistics for the variables you are planning to use.


# Data Wrangling and Transformation

What data cleaning did you have to do?
How did you wrangle the data?
Are you deciding to exclude any observations? If so, why?
Did you have to create any new variables from existing variables? If so, how and why?


Add a codebook – a description of all variables you are using, including ones you are creating for this
project – to the README.md page of the data/ folder of your repo.

# Data Visualization
```{r}
#| label: viz-one
#| fig-cap: This figure shows 
#| echo: true

# figure

```

```{r}
#| label: viz-two
#| fig-cap: This figure shows xyz
#| echo: true

# figure

```

```{r}
#| label: viz-three
#| fig-cap: This figure showssss
#| echo: true

# figure

```

```{r}
#| label: viz-four
#| fig-cap: This figure shows abc
#| echo: true

# figure

```
