---
title: "EDA"
subtitle: "4/16 Assignment"
format: 
  html:
    page-layout: full
    theme: cosmo
    toc: true
    toc-title: "Contents"
    embed-resources: true
    standalone: true
editor: 
  markdown: 
    wrap: 72
---

# Exploring the data

We are looking at bird sightings at sea, which comes from TidyTuesday at https://github.com/rfordatascience/tidytuesday/tree/main/data/2026/2026-04-14. Each observation communicates the data that an observer witnessed during a 10-minute period at a port.

```{r}
#| label: loading in the data
#| echo: true
#| warning: false
#| message: false 

tuesdata <- tidytuesdayR::tt_load('2026-04-14')

beaufort_scale <- tuesdata$beaufort_scale
birds <- tuesdata$birds
sea_states <- tuesdata$sea_states
ships <- tuesdata$ships

```

```{r}
#| label: load packages
#| echo: true 
#| warning: false
#| message: false

library(tidyverse) # for general purposes and ggplot 
library(gtsummary) # for nice tables
library(maps) # for plotting latitude and longitude 
```

```{r}
#| label: merging dfs
#| echo: true

# merge ship and sea data 
ships <- ships %>%
  left_join(beaufort_scale, by = "wind_speed_class") %>%
  left_join(sea_states, by = "sea_state_class") %>%
  mutate(date = as.Date(date),
         year = year(date),
         month = month(date))

# merge birds with ship and sea data
df <- birds %>%
  left_join(ships, by = "record_id")
```

```{r}
#| label: managing missing data
#| include: false

# total number of missing values
sum(is.na(df))
# proportion of missing values in entire dataset
mean(is.na(df))

#want to see if certain cols have a lot missing
missing_summary_raw_df <- df %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_raw_df

# 
# common_name (or group species!)
# protocol_type
# duration_minutes
# effort_distance_km
# number_observers
# hour
# season
# location (stat
#           
#clean the data. did it this way because there is a missing data problem both by row and by column, and clearing either individually would leave us with no data 

df_clean <- df %>%
  filter(!is.na(count)) %>% # these are variables we definitely want to use, so filter out observations that are missing these. species common name has the fewest missing obs of the three different types of species name columns. doing them in separate lines because the or command wasn't working correctly for some reason
  filter(!is.na(latitude)) %>%
  filter(!is.na(longitude)) %>%
  filter(!is.na(year)) %>%
  filter(!is.na(month)) %>%
  filter(!is.na(species_common_name)) %>% 
  select(where(~ mean(is.na(.)) <= 0.6)) %>% 
  filter(rowMeans(is.na(.)) <= 0.6) %>%
  select(where(~ mean(is.na(.)) <= 0.2)) %>%
  filter(rowMeans(is.na(.)) <= 0.0) 

# see where we're at after removing some things 
missing_summary_cleaned_df <- df_clean %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "column",
               values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

missing_summary_cleaned_df

# total number of missing values
sum(is.na(df_clean))
# proportion of missing values in entire dataset
mean(is.na(df_clean))

```

```{r}
#| label: making some new variables
#| echo: true

# species column has a LOT of species that are observed very few times. going to make a new variable that collapses rare ones into one "other" column
df_clean <- df_clean %>%
  group_by(species_common_name) %>%
  mutate(species_count = n()) %>%
  ungroup() %>%
  mutate(species_simplified = case_when(
    species_count/nrow(df_clean) >= 0.01 ~ species_common_name,
    TRUE ~ "other"
  )) %>%
  select(-species_count)


# this still has a lot of unique birds. it may be better to do family level
df_clean <- df_clean %>%
  mutate(species_family = case_when(
    str_detect(species_common_name, "albatross|mollymawk") ~ "Albatross",
    str_detect(species_common_name, "petrel|shearwater|prion|fulmar|diving-petrel") ~ "Petrel/Shearwater/Prion",
    str_detect(species_common_name, "gull|tern|noddy") ~ "Gull/Tern",
    str_detect(species_common_name, "skua|jaeger") ~ "Skua/Jaeger",
    str_detect(species_common_name, "penguin") ~ "Penguin",
    str_detect(species_common_name, "gannet|booby") ~ "Gannet/Booby",
    str_detect(species_common_name, "cormorant|shag") ~ "Cormorant/Shag",
    str_detect(species_common_name, "frigatebird|tropicbird") ~ "Frigatebird/Tropicbird",
    TRUE ~ "Unidentified/Other"
  )) 

```

```{r}
#| label: summary stats table 
#| echo: true

# factoring species so that unidentified comes last 
df_clean <- df_clean %>%
  mutate(species_family = factor(species_family,
                                levels = c(
                                  "Albatross",
                                  "Petrel/Shearwater/Prion",
                                  "Gull/Tern",
                                  "Skua/Jaeger",
                                  "Penguin",
                                  "Gannet/Booby",
                                  "Cormorant/Shag",
                                  "Frigatebird/Tropicbird",
                                  "Unidentified/Other"
                                )))

# print table. selecting our core variables, which are the ones that we are investigating in EDA 
df_clean %>%
  select(
    species_family,
    count,
    season,
    cloud_cover, precipitation,
    feeding, flying_past, following_ship,
    sea_state_description,
    longitude,
    latitude
  ) %>%
  mutate(
    season = str_to_title(season),
    precipitation = str_to_title(precipitation),
    cloud_cover = str_to_title(cloud_cover),
    sea_state_description = str_to_title(sea_state_description)
  ) %>%
  rename(
         "Bird Family" = species_family,
         "Bird count" = count,
         "Number of birds feeding" = feeding,
         "Number of birds flying past" = flying_past,
         "Number of birds following ships" = following_ship,
         "Cloud cover" = cloud_cover,
         "Season" = season,
         "Precipitation" = precipitation,
         "Sea state" = sea_state_description,
         "Longitude" = longitude,
         "Latitude" = latitude
         ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
    as_gt() %>%
  gt::tab_caption("Table 1. Summary of bird observations")

#write csv of cleaned data
write_csv(df_clean, "birds_clean.csv")
```

Our key outcome variable currently is bird count, which measures how many total birds an observer witnessed during an observation session. We are secondarily interested in looking at counts of birds doing specific activities, such as feeding, flying past, and following ships. Our key explanatory variables are season, longitude, latitude, precipitation, cloud cover, and sea state. We expect that the number of birds will vary dramatically depending on the season and location, but that poor weather will also affect bird count at a more granular level.

# Data Wrangling and Transformation

We wrangled the data via the base R object `tt_load`, which allows us to directly load TidyTuesday data from Github. We merged the separate data frames by the TidyTuesday-recommend identifier columns recommended to form a data frame with `r dim(df)[1]` observations of `r dim(df)[2]` variables.

As for data cleaning, we chose to exclude some observations and variables with missing data to ensure that our exploratory assessments were not distorted by data that would need to be removed for modeling purposes. We first removed any observations that were missing our core variables: bird count, species common name, latitude, longitude, year, and month. Next, we performed an iterative process of removing columns and observations with missing data; we removed variables with more than 60% of observations missing, then observations with more than 60% of variables missing, then variables with more than 20% of observations missing, and, finally, observations with any missing data. This iterative process allowed us to retain `r dim(df_clean)[1]` observations and `r dim(df_clean)[2]` variables.

We created a new variable that categorized birds by family rather than species because most individual species were observed very few times, but overall families were observed frequently.

::: {.callout collapse="false"}
## Codebook

| Variable | Type | Description |
|------------------|------------------|-------------------------------------|
| `species_family` | character | Bird family, created by grouping species |
| `count` | integer | Number of birds counted in the observation |
| `feeding` | integer | Number of birds observed feeding |
| `flying_past` | integer | Number of birds observed flying past the ship |
| `following_ship` | integer | Number of birds observed following the ship |
| `cloud_cover` | ordered | "clear", "partially cloudy" or "overcast" |
| `season` | ordered | Southern hemisphere season |
| `precipitation` | factor | Precipitation type, including levels of rain, snow, and fog |
| `sea_state_description` | ordered | Description of sea state, from "calm, glassy" to "very rough" |
| `longitude` | double | Decimal longitude |
| `latitude` | double | Decimal latitude (negative indicates Southern hemisphere) |
:::

# Data Visualization

```{r}
#| label: bar plot (not using)
#| include: false
#| echo: false

# grab summary so we can plot iqr
df_summary <- df_clean %>%
  distinct(record_id, .keep_all = TRUE) %>% # so we don't count one session multiple times
  group_by(season) %>%
  summarise(
    mean_count = mean(count, na.rm = TRUE),
    q1 = quantile(count, 0.25, na.rm = TRUE),
    q3 = quantile(count, 0.75, na.rm = TRUE)
  )

ggplot(df_summary, aes(x = season, y = mean_count)) +
  geom_col() +
  geom_errorbar(aes(ymin = q1, ymax = q3), width = 0.2) +
  labs(title = "Mean Number of Birds Observed by Season with IQR",
       x = "Season",
       y = "Mean Number of Birds") +
  theme_minimal()

```

```{r}
#| label: viz-one
#| fig-cap: Boxplot of log-transformed number of birds observed by season. Log-transformation was used due to the data's extreme right skew.
#| echo: true

# take log count of birds 
df_clean <- df_clean %>%
  mutate(log_count = log10(count + 1))

# boxplot of log count by season
season_box <- ggplot(df_clean, aes(x = season, y = log_count, fill = season)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Log Bird Counts by Season",
       x = "Season",
       y = "Log Bird Count") +
  theme(legend.position = "none")

print(season_box)

```

```{r}
#| label: viz-two
#| fig-cap: Map of latitude and longitude clusters based on k-means clustering with k=3. 
#| echo: true

### perform k-means clustering 
set.seed(244) # for reproducibility 

## first, use the elbow method to find the optimal 'k'
# set possible values of k (picked 10 randomly)
k_values <- 1:10

# create a function to compute the Total Within-Cluster Sum of Squares (WCSS)
# for a given k
get_wcss <- function(k) {
  kmeans_model <- kmeans(df_clean %>% select(longitude, latitude), centers = k,
                         nstart = 25)
  return(kmeans_model$tot.withinss)
}

# apply function to the range of k values
set.seed(244) # set seed for reproducibility
wcss_values <- map_dbl(k_values, get_wcss)

# store the results in a new dataframe
elbow_df <- tibble(
  k = k_values,
  wcss = wcss_values
)

# plot the elbow curve plot. commented this out because we don't
# want it to print for our EDA
# ggplot(elbow_df, aes(x = k, y = wcss)) +
#   geom_line() +
#   geom_point() +
#   scale_x_continuous(breaks = k_values) + # force x-axis to show whole numbers
#   theme_minimal() +
#   labs(title = "Elbow Method for Optimal 'k'",
#        x = "Number of Clusters (k)",
#        y = "Total Within-Cluster Sum of Squares (WCSS)")

# it looks like three clusters is probably satisfactory

## now, do the k-means clustering
kmeans_result <- kmeans(df_clean %>% select(longitude, latitude),
                        centers = 3, nstart = 25)

# add the cluster assignments back to the dataset
df_clean$cluster <- as.factor(kmeans_result$cluster)

# visualize the clusters on a map via maps package 
world_map <- map_data("world")

# now plot the cluster results 
ggplot() +
  # draw the world map 
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
               fill = "lightgray", color = "white") +
  # plot the clustered observations
  geom_point(data = df_clean, aes(x = longitude, y = latitude, color = cluster), 
             alpha = 0.6, size = 1) +
  coord_quickmap(xlim = c(40, 200), ylim = c(-90, 0)) + # zooming map in to
  # just australia-ish area
  theme_minimal() +
  labs(title = "Bird Observations Clustered by Location",
       subtitle = "K-Means Clustering (k=3)",
       x = "Longitude", 
       y = "Latitude",
       color = "Cluster") +
  theme(legend.position = "bottom")

```

```{r}
#| label: viz-three
#| fig-cap: Frequency of observed bird activities across all observation sessions. 
#| echo: true

# defining activity columns
activity_cols <- c("feeding", "sitting_on_water", "sitting_on_ice", "sitting_on_ship",
                   "in_hand", "flying_past", "accompanying", "following_ship", "naturally_feeding")

# reshape data from wide to long format to summarize activities, and make the activities more nicely formatted for the plot 
df_activities <- df_clean %>%
  select(all_of(activity_cols)) %>%
  pivot_longer(cols = everything(), names_to = "activity_type", values_to = "is_doing") %>%
  # filter only the TRUE observations
  filter(is_doing == TRUE) %>%
  mutate(
    activity_type = str_replace_all(activity_type, "_", " "), # replace underscores with spaces
    activity_type = str_to_title(activity_type)               # capitalize first letters
  )

# plot the frequency of each activity
ggplot(df_activities, aes(x = fct_infreq(activity_type))) +
  geom_bar(fill = "red4", color = "black", alpha = 0.8) +
  coord_flip() + # flip coordinates for easier reading of labels
  theme_minimal() +
  labs(title = "Frequency of Observed Bird Activities",
       x = "Activity Type",
       y = "Total Occurrences")

```

```{r}
#| label: viz-four
#| echo: true
#| fig-cap: Boxplot of log bird count by sea state.

df_sea_state <- df_clean %>% 
  mutate(
    # capitalize the first letter of each word
    sea_state_description = str_to_title(sea_state_description) 
  )

# boxplot of log count by sea state
sea_state_box <- ggplot(df_sea_state, 
             aes(x = reorder(sea_state_description, log_count, FUN = median), y = log_count)) +
  geom_boxplot(fill = "red4", alpha = 0.7) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Bird Counts Relative to Sea State",
       x = "Sea State",
       y = "Log Bird Count")

print(sea_state_box)
```
