---
title: "Further Exploration and Model"
subtitle: "4/30 Presentation"
format: 
  html:
    page-layout: full
    theme: cosmo
    toc: true
    toc-title: "Contents"
    embed-resources: true
    standalone: true
editor: 
  markdown: 
    wrap: 72
---

# Further Exploration

We noticed from EDA that the `count` variable was very right-skewed, and had a large range including regions with very sparse data. To examine this further, and to potentially determine if there was a range of the `count` data that could be used for prediction, we decided to look at the range from a count of 10 to the mean (51.9) plus one standard deviation (1616.5), which is 1668.4 birds. 
```{r}
#| label: loading in the data
#| echo: true
#| warning: false
#| message: false 
library(tidyverse)
library(caret) # will be needed for CV later
library(pROC) # will be needed for logistic regression models
library(rpart) #for decision tree
library(rpart.plot) #to plot decision tree

#read in the cleaned data
data <- read_csv("birds_clean.csv")

# Plot the censored data
data_smallcount <- data %>%
  filter(count > 10, count < 1668.4)

ggplot(data_smallcount, aes(x = count)) + 
  geom_histogram(binwidth = 10) +
  labs(
    title = "Bird Count: Censored (10 < count < 1668.4)",
    x = "Bird Count"
  ) +
  theme_minimal()


##3 aggregate records so total_birds is the total number of birds recorded in a given session, since currently "count" is total numbers of a specific species of birds recorded in a given session 
session_data <- data %>%
  group_by(record_id) %>%
  summarize(
    # calculate total birds seen during the entire observation session
    total_birds = sum(count, na.rm = TRUE),
    
    # keep the spatial and environmental variables for the session
    latitude = first(latitude),
    longitude = first(longitude),
    sea_state_description = first(sea_state_description),
    wind_description = first(wind_description),
    precipitation = first(precipitation),
    season = first(season),
    cloud_cover = first(cloud_cover),
    .groups = 'drop'
  ) %>%
  mutate( # also make birds interval 
    bird_interval_four = cut(total_birds,  # by four
                        breaks = c(-Inf, 10, 50, 100, Inf), 
                        labels = c("0-10", "11-50", "51-100", "100+"),
                        ordered_result = TRUE),
    bird_interval_two = cut(total_birds,  # by two (binary)
                        breaks = c(-Inf, 50, Inf), 
                        labels = c("0-50", "51+"),
                        ordered_result = TRUE)
  )

#plot  aggregates
ggplot(session_data %>% filter(total_birds < 1000), aes(x = as.numeric(total_birds))) + 
  geom_histogram(binwidth = 10) +
  labs(
    title = "Bird Count",
    x = "Bird Count"
  ) +
  theme_minimal()

```

Looking at this much smaller region of the `count` data compared to the whole dataset, this is much more interpretable, but we can still see the issue of a heavy right-skew and regions with sparse data. Since there were so many observations with no birds, we decided to make a new variable that transformed the bird count variable into a binary variable. Then, we could make a Logistic Regression model to predict the appearance of birds. We decided to transform the total number of birds variable in two ways: the first being `count` is greater than 1 or `count` is less than or equal to 1, and the second being `total_birds` is 0-50, or 51+. The first model uses the `df_logistic` dataset, and the second model uses the aggregated `session_data` set. 

```{r}
#| label: bird appearance binary variable
#| echo: false
#| warning: false 
#| message: false

# create a new variable that is a binary of whether or not there is a bird
df_logistic <- data %>%
  mutate(birds = case_when(count > 1 ~ "X1",
                           count <= 1 ~ "X0"))

# filter variables
df_logistic <- df_logistic %>%
    select(
    species_family,
    birds,
    season,
    cloud_cover, precipitation,
    feeding, flying_past, following_ship,
    sea_state_description,
    longitude,
    latitude
  ) %>%
  mutate(cloud_cover = factor(cloud_cover, levels = c(
    "clear", "partially cloudy", "overcast"
  ))) %>%
  mutate(sea_state_description = factor(sea_state_description, levels = c(
    "calm, rippled", "smooth", "slight", "moderate", "rough", "very rough"
  ))) %>%
  mutate(birds = factor(birds))

#summary(df_logistic$birds)
# set seed
set.seed(244)
# split into test/ train for CV
test1_id <- sample(1:nrow(df_logistic), size= nrow(df_logistic) * .25)
train1 <- df_logistic[-test1_id,]
test1 <- df_logistic[test1_id,]
# set up control for classification
ctrl <- trainControl(method = "cv",
                     number = 10,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)
# train model
lr_1 = train(
  form = birds ~ season + cloud_cover + sea_state_description +
    precipitation + species_family + longitude + latitude +
    feeding + flying_past + following_ship,
  data = train1,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl)
lr_1
# test model
test1_pred <- predict(lr_1, newdata = test1)
test1_prob <- predict(lr_1, newdata=test1, type= "prob")
# make confusion matrix
cm1 <- confusionMatrix(data = test1_pred, reference=test1$birds)
cm1
# AUC
roc1 <- roc(test1$birds, test1_prob$X0)
auc(roc1)
# plot curve
ggroc(roc1, legacy.axes = TRUE) + #set axes to 1-specificity vs sensitivity
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") + #AUC = 0.5
  labs(
    title = "ROC Curve of Model 1",
    x= "1 - specificity",
    y= "sensitivity") + 
  theme_minimal()
```
```{r}
#| label: aggregated data model
#| echo: false
#| message: false
#| warning: false

# code the bird_interval_two variable as a factor variable for logistic regression
lr2_data <- session_data %>% 
  mutate(birds = case_when(bird_interval_two == "0-50" ~ "X0",
                           bird_interval_two == "51+" ~ "X1")) %>%
  select(birds, latitude, longitude,
         sea_state_description, 
         wind_description, precipitation, season, cloud_cover) %>%
  mutate(birds = factor(birds))
#note: aggregated data is missing a lot of potential predictor variables

# set seed
set.seed(244)
# split into test/ train for CV
test2_id <- sample(1:nrow(lr2_data), size= nrow(lr2_data) * .25)
train2 <- lr2_data[-test2_id,]
test2 <- lr2_data[test2_id,]
# set up control for classification
ctrl <- trainControl(method = "cv",
                     number = 10,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)
# train model
lr_2 = train(
  form = birds ~ .,
  data = train2,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl)
lr_2
# test model
test2_pred <- predict(lr_2, newdata = test2)
test2_prob <- predict(lr_2, newdata=test2, type= "prob")
# make confusion matrix
cm2 <- confusionMatrix(data = test2_pred, reference=test2$birds)
cm2
# AUC
roc2 <- roc(test2$birds, test2_prob$X1)
auc(roc2)
# plot curve
ggroc(roc2, legacy.axes = TRUE) + #set axes to 1-specificity vs sensitivity
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") + #AUC = 0.5
  labs(
    title = "ROC Curve of Model 2",
    x= "1 - specificity",
    y= "sensitivity") + 
  theme_minimal()
```
Based on these two logistic regression models, it appears that the model that separates the `birds` variable into `count` is greater than 1 and `count` is less than or equal to 1 (Model 1) performs slightly better (AUC_1 = 0.6384 vs AUC_2 = 0.5815). This indicates that Model 1 has a slightly better discrimination ability than Model 2, though neither is excellent. 

To further examine our data, we decided to run a Decision Tree to classify the total birds into smaller intervals. 

```{r}
#| label: decision tree
#| echo: false
#| warning: false
#| message: false

# filter data for tree-making
tree_data <- data %>%
  mutate(birds_four = case_when(count <= 10 ~ "0-10",
                                count <= 50 ~ "11-50",
                                count <= 100 ~ "51-100",
                                count > 100 ~ "100+")) %>%
  mutate(birds_four = factor(birds_four)) %>%
  select(birds_four,
    season,
    cloud_cover, precipitation,
    sea_state_description,
    longitude,
    latitude,
    wind_description
  ) %>% 
  mutate(cloud_cover = factor(cloud_cover, levels = c(
    "clear", "partially cloudy", "overcast"
  ))) %>%
  mutate(sea_state_description = factor(sea_state_description, levels = c(
    "calm, rippled", "smooth", "slight", "moderate", "rough", "very rough"
  ))) %>%
  mutate(wind_description = factor(wind_description))

# set seed
set.seed(244)
# test/train sets
test3_id <- sample(1:nrow(tree_data), size= nrow(tree_data)/2)
train3 <- tree_data[test3_id,]
test3 <- tree_data[-test3_id,]
# grow decision tree
bird_tree <- rpart(birds_four ~ ., data=test3, method="class", cp=0)
#plot tree
rpart.plot(bird_tree, box.palette="Blues")
#test error
test3_err <- predict(bird_tree, newdata=test3, type="class")
table(test3_err, test3$birds_four)
(924/11691)
```

The test error for the Decision Tree is 0.079, which is very low. Based on the confusion matrix, the tree is good at classifying the more common categories of 0-10 birds and 11-50 birds, which are also the two categories that indicate the smallest number of birds. This makes sense, as around half of the data reports numbers of birds between 0-50 and half 50+. 